Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-04-23 paper.tex
Error of the rotation

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix for coordinates of points in the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample makes the direction of
anisotropy parallel to one of the axes, so anisotropy can be better represented in
the correlation function \cref{fig:aniso}.
\begin{figure*}[!htp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.3\linewidth, frame]{images/aniso.png} &
    \includegraphics[width=0.3\linewidth, frame]{images/aniso-rot.png} & \\
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png} \\
  \end{tabular}
  \caption{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is elongated along the axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to the algorithm described in
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

>>>>

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix for coordinates of points in the
phase of interest:
\begin{equation}
  C_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $C$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample makes the direction of
anisotropy parallel to one of the axes, so anisotropy can be better represented in
the correlation function \cref{fig:aniso}. In our implementation we use rotation
with nearest-neighbor interpolcation which affects quality of computations as
discussed in \cref{sec:rotqua}.
\begin{figure*}[!htp]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.3\linewidth, frame]{images/aniso.png} &
    \includegraphics[width=0.3\linewidth, frame]{images/aniso-rot.png} & \\
    \includegraphics[width=0.3\linewidth]{images/aniso-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png} &
    \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png} \\
  \end{tabular}
  \caption{Behaviour of our algorithm for calculation of the $S_3$ function
    when dealing with highly anisotropic media. The pattern \cref{fig:pattern}
    which we use in our algorithm is most sensitive to anisotropy which is
    directed along one of the axes. \\
    From top down, left to right: an original image of an anisotropic pore; a
    rotated image of the pore, so it is elongated along the axis; $S_3$ function
    for the original pore; $S_3$ function for the rotated pore with a vector
    $(0.92, -0.4)$ being a direction of elongation of the pore; $S_3$ function
    averaged across rotations similar to the algorithm described in
    \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

\section{Influence of sample rotation on quality of computations}
\label{sec:rotqua}
\begin{figure*}[tp]
  \centering
  \subfigure[A sample of value noise]{
    \includegraphics[width=0.4\linewidth]{images/noise.png}
    \label{fig:roterror-noise}}
  \hfill
  \subfigure[Relative error of computation]{
    \includegraphics[width=0.4\linewidth]{images/roterror.png}
    \label{fig:roterror-error}}
  \caption[]{A plot of dependency of maximal pointwise relative error
    \cref{eq:relerr} in computation of correlation functions introduced by a
    rotation of the sample \cref{fig:roterror-noise} from the rotation
    angle. The rotation is performed around $Z$ axis counter-clockwise using
    nearest-neighbor interpolation.}
  \label{fig:roterror}
\end{figure*}
Our edge extraction filter works only for binary samples, therefore it is
natural to require that a rotated image remains binary. The simpliest way to do
so is to use nearest-neighbor interpolation when sampling subvoxel data during
rotation. This results only in minimal precision loss in computation of
correlation functions.

Speculatively, the loss in precision should be minimal for three-point and
cluster correlation functions, because they are calculated
``in the whole volume'' of the sample. The biggest loss should be for $F_{sss}$
function because it takes into account only a small volume of the sample which
lies closely to the interface between phases and this volume is the only volume
affected by the rotation.

To estimate the error introduced by rotation, consider a
$300 \times 300 \times 300$ sample of value noise \cref{fig:roterror-noise}
which is defined as $f(x, y, z) > T$ where $f$ is a value noise function and $T$
is a threshold. We produce images of the noise rotated around $Z$ axis by $\phi$
radians by discretizing an inequation $f(p(x'), p(y'), p(z')) > T$ where
\begin{equation*}
  \begin{aligned}
    x' &= x \cos \phi + y \sin \phi \\
    y' &= y \cos \phi - x \sin \phi \\
    z' &= z
  \end{aligned}
\end{equation*}
and
\begin{equation*}
  \begin{aligned}
    p(x) &= \left\{
    \begin{array}{ll}
      x'     & \quad x' < 1 \\
      2 - x' & \quad \text{otherwise}
    \end{array}
    \right. \\
    x ' &= x \mod 2
  \end{aligned}
\end{equation*}
Therefore we assume the noise to be periodic with the period equal to
$2$ and the interface between phases is continuous. Discretizing for different
values of $\phi \in [0, \pi/4]$ we obtain the first dataset which contains the
best possible discretization of the rotated noise.

The second dataset is constructed by discretizing the noise with $\phi = 0$ and
then rotating the discretized image using nearest-neighbor interpolation.

We compute $S_3$ and $F_{sss}$ functions in periodic mode for the both datasets
and calculate pointwisely a relative error introduced by the rotation as
\begin{equation}
  \varepsilon = \frac{|F_1 - F_2|}{F_1}
  \label{eq:relerr}
\end{equation}
where $F_1$ is a correlation function for a sample from the first dataset and
$F_2$ is a correlation function for a sample from the second dataset. The we
calculate the maximal relative error for each $\phi$ as the maximum of all
pointwise errors for a given rotation angle.

Plots of maximal relative errors for functions $S_3$ and $F_{sss}$ are on
\cref{fig:roterror-error}. For the surface-surface-surface function we omit
values $F_{sss}(s_1, s_2)$ if either of its arguments is less than 10 voxels
because this function is undefined for either $s_1$ or $s_2$ being equal to zero
and our edge detection filter gives at least 7 voxel transition zone where the
correlation function is affected by this indeterminacy. The maximal error for
all $\phi$'s are 0.6\% for $S_3$ and 4.4\% for $F_{sss}$.
